W(ha/il)t's the (wo/ba)rd?

### 2022-04-24 paper_5.rmd
updates

---
title: "W(ha/il)t's the (wo/ba)rd?" # TODO
subtitle: "A cross-sectional analysis of English word pairs that satisfy the properties of the Split Decisions word puzzle"
author: "Oliver Daniel"
abstract: >
  From three English wordlist corpora, pairs of words are generated which might appear in the Split Decisions word puzzle as published by the New York Times, i.e., at least 5 characters in length and differing by exactly two consecutive characters. The corpora are contrasted among themselves, and then the lexical and phonological properties of those words which form pairs – and those which do not – are analyzed for causative insights. Such insights will be alluded to here once discovered.

date: "April 6, 2022"
output:
  bookdown::pdf_document2:
    toc: true
    number_sections: false
knit: (function(input, ...) {
    rmarkdown::render(
    input,
    output_file = "out/paper_5.pdf",
    envir = globalenv(),
    ...
    )
  })
bibliography: in/references.bib
link-citations: true
---
\\par \\textbf{Keywords}: crosswords, split decisions, computational linguistics

Code available on [GitHub (link)](https://github.com/oliver-daniel/inf_313_final_paper).
```{r setup, echo=FALSE}
knitr::opts_chunk$set(
    echo = FALSE, warning = FALSE,
    out.width = "100%", out.align = "center", dir
)
options(knitr.graphics.error = FALSE)
```

@@

## Introduction
A few times a year, the New York Times games department releases a new issue of Split Decisions, a crossword-like word puzzle. Similar to crosswords, the game board consists of boxes arranged in sequence, one for each letter of a word running either left-to-right or top-to-bottom, intersecting with other words at right angles. However, instead of cryptic descriptions of correct answers, Split Decisions solvers are only given pairs of two-letter substrings for each word, either of which serves to complete a valid English word. For example, given the following diagram:

@@

A correct fill for the remaining boxes might be __EASE, creating both PLEASE and GREASE. For the remainder of this paper, such a pair of words will be referred to as a *split pair*. Conversely, an incorrect fill might be __IERS, which produces the valid word PLIERS and the invalid word \*GRIERS[^1]. Note that each combination of clues (hereon a *split*) might accept several different fills for the common substring, e.g., __UNGE (PLUNGE, GRUNGE); however, as with crosswords, only one solution will also satisfy all the crossing constraints produced by the rest of the puzzle.

@@

```{r}
corpus_stats |>
    left_join(split_x_counts, by = "corpus") |>
    left_join(split_y_counts, by = "corpus") |>
    left_join(split_pair_counts, by = "corpus") |>
    mutate(
        split_x = paste0(split_x, " (", split_x_count, ")"),
        split_y = paste0(split_y, " (", split_y_count, ")"),
        split_pair = paste0(split_pair, " (", split_pair_count, ")")
    ) |>
    select(corpus, all_words, valid_words, pairs_count, split_x, split_y, split_pair) |>
    kable(
        booktabs = TRUE,
        col.names = c("Corpus", "Total",
            "Valid", "Total", "First split",
            "Second split", "Split pair")
    ) |>
    kable_classic() |>
    add_header_above(c(" " = 1, "Words" = 2, "Pairs" = 1, "Most frequent (occurrences)" = 3))
```

@@

<!-- Footnotes -->
[^1]: The asterisk prefix will be used throughout this paper to indicate incorrect/invalid words.

>>>>

---
title: "W(ha/il)t's the (wo/ba)rd?"
thanks: >
  "What's the word?" or "Wilt's the bard?" (nonsense).
subtitle: "A cross-sectional analysis of English word pairs that satisfy the properties of the Split Decisions word puzzle"
author: "Oliver Daniel"
abstract: >
  From three English wordlist corpora, pairs of words are generated which might appear in the Split Decisions word puzzle as published by the New York Times, i.e., at least 5 characters in length and differing by exactly two consecutive characters. The corpora are contrasted among themselves, and then the lexical and phonological properties of those words which form pairs – and those which do not – are analyzed for causative insights. Such insights will be alluded to here once discovered.
date: "April 27, 2022"
output:
  bookdown::pdf_document2:
    toc: true
    number_sections: false
knit: (function(input, ...) {
    rmarkdown::render(
    input,
    output_file = "out/paper_5.pdf",
    envir = globalenv(),
    ...
    )
  })
geometry: "margin=1.5in"
bibliography: in/references.bib
link-citations: true
---
\par \textbf{Keywords}: crosswords, split decisions, computational linguistics

Code available on [GitHub (link)](https://github.com/oliver-daniel/inf_313_final_paper).
```{r setup, echo=FALSE}
knitr::opts_chunk$set(
    echo = FALSE, warning = FALSE, cache = TRUE,
    out.width = "100%", out.align = "center", dir
)
options(knitr.graphics.error = FALSE)
```

@@

\tableofcontents

## Introduction
A few times a year, the New York Times games department releases a new issue of a word puzzle called Split Decisions. Similar to crosswords, the game board consists of boxes arranged in sequence, one for each letter of a word running either left-to-right or top-to-bottom, intersecting with other words at right angles. However, instead of cryptic descriptions of correct answers, Split Decisions solvers are only given pairs of two-letter substrings for each word, either of which serves to complete a valid English word. For example, given the following diagram:

@@

A correct fill for the remaining boxes might be __EASE, creating both PLEASE and GREASE. For the remainder of this paper, such a pair of words will be referred to as a *word pair*, sharing a *common substring*[^2] and differing by a *split pair*. Conversely, an incorrect fill might be __IERS, which produces the valid word PLIERS and the invalid word \*GRIERS[^1]. Note that each clued split pair might accept several different fills for the common substring, e.g., __UNGE (PLUNGE, GRUNGE); however, as with crosswords, only one solution will also satisfy all the crossing constraints produced by the rest of the puzzle.

@@

```{r}
corpus_stats |>
    left_join(split_x_counts, by = "corpus") |>
    left_join(split_y_counts, by = "corpus") |>
    left_join(split_pair_counts, by = "corpus") |>
    mutate(
        split_x = paste0(split_x, " (", split_x_count, ")"),
        split_y = paste0(split_y, " (", split_y_count, ")"),
        split_pair = paste0(split_pair, " (", split_pair_count, ")")
    ) |>
    select(corpus, all_words,
    valid_words, pairs_count, split_x,
    split_y, split_pair) |>
    kable(
        booktabs = F,
        col.names = c("Corpus", "Total",
            "Valid", "Total", "First split",
            "Second split", "Split pair")
    ) |>
    kable_classic(latex_options = "hold_position") |>
    add_header_above(c(" " = 1, "Words" = 2, "Word pairs" = 1, "Most frequent (occurrences)" = 3))
```

@@

<!-- Footnotes -->
[^1]: The asterisk prefix will be used throughout this paper to indicate incorrect/invalid words.
[^2]: Or *fill*: this part of the puzzle is filled in by the player rather than the constructor.

### 2022-04-24 scripts/exploration.rmd
updates

>>>>

```{r}
# TODO: better gridline color
corpus_splits |>
  ggplot(aes(
    x = substr(splits, 1, 1),
    y = substr(splits, 2, 2)
  )) +
  geom_bin2d() +
  geom_bin2d(data = transform(corpus_splits, corpus = "total")) +
  coord_fixed() +
  scale_fill_continuous(type = "viridis") +
  facet_wrap(~corpus, ncol = 2)
```

```{r}
dat.all() |>
  filter(length <= 15) |>
  ggplot(aes(
    x = factor(length),
    y = start,
    color = corpus
  )) +
  geom_violin(scale = "width") +
  facet_wrap(~corpus, dir = "v") +
  scale_y_continuous(breaks = c(2, 4, 6, 8, 10, 12, 14, 16)) +
  theme(
    legend.position = "none"
  )
```

```{r}
# Interested in different combinations of "templates" --
# how many length + split + posn combos are there?
templates <- splits_by_type |>
  distinct(x, y, .keep_all = T) |>
  add_count(length, split_x, split_y, start, type)

templates |>
  group_by(split_x, split_y, type) |>
  arrange(desc(n)) |>
  View()
```

```{r}
# Similarly, what about coming into and out of splits?

splits_by_type |>
  filter(type == "prefix") |>
  mutate(common_first = substr(x, end + 1, end + 1)) |>
  count(common_first, split_x, split_y) |>
  arrange(desc(n))
# shows coarticulation in action!!! "im" for p, in/un for t/s...

splits_by_type |>
  filter(type == "suffix") |>
  mutate(common_last = substr(x, start - 1, start - 1)) |>
  count(common_last, split_x, split_y) |>
  arrange(desc(n))
# lots of i + declension... what else?

splits_by_type |>
  filter(type == "suffix") |>
  mutate(common_last = substr(x, start - 1, start - 1)) |>
  filter(common_last != "i") |>
  count(common_last, split_x, split_y) |>
  arrange(desc(n))
# t, s, a, e, n here with other kinds of declension
# (ed + or, nt + te)... if I were in the mood, I'd
# look for etymology too (maybe these are Latinate stems?)
```

```{r}
# vowels going into vowels
splits_by_type |>
  mutate(common_last = substr(x, start - 1, start - 1)) |>
  filter(str_detect(split_x, "^[aeiou]") | str_detect(split_y, "^[aeiou]")) |>
  filter(common_last %in% c("a", "e", "i", "o", "u")) |>
  count(common_last, split_x, split_y, type) |>
  arrange(desc(n)) #|> View()
  # once scrolling past the domination of i, some interesting effects,
  # esp with vowels that can double (oo, ee) -- Great Vowel Shift spellings?

  splits_by_type |>
  filter(substr(x, start - 1, start - 1) == "o" &
        (substr(split_x, 1, 1) == "o" |
         substr(split_y, 1, 1) == "o"))# |> View()
  # as expected, many splits with "o_" have /u/ represented by the <oo>, but
  # some are even tricker due to other consonant interactions (floor/flown)

  splits_by_type |>
  mutate(
    common_last = substr(x, start - 1, start - 1),
    split_x_first = substr(split_x, 1, 1),
    split_y_first = substr(split_y, 1, 1)
  ) |>
  filter(common_last %in% c("o", "e") &
        (common_last == split_x_first | common_last == split_y_first)) |>
  filter(!if_else(
    common_last == split_x_first,
    split_y_first,
    split_x_first
  ) %in% c("a", "e", "i", "o", "u")) |> View()

  # Also introducing some neat schwa/lenition effects:
  # "belay/beefy", etc.
  # time to generalize this
```

```{r}
# get all rows where the:
# a) last letters of the common, and
# b) first letter of either split
# are those specified.
withLeftBoundary <- function(data, last_common_letters, first_split_letters) {
  data |>
  mutate(
    common_last = substr(x, start - 1, start - 1),
    split_x_first = substr(split_x, 1, 1),
    split_y_first = substr(split_y, 1, 1)
  ) |>
  rowwise() |>
  filter(common_last %in% last_common_letters &
         length(intersect(c(split_x_first, split_y_first),
         first_split_letters)) > 0)
}
```
